---
title: 'Rating Prediction for MovieLens'
author: "José Alejandro Luna Matos"
date: "September 6, 2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE , echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```


# MovieLens Project  

The goal of the project  is to build a recommendation system for movies using machine learning in R . 
In order to get that , the Data will be explored , transform and exploited.



```{r ,libraries,include = FALSE, echo=TRUE}
# Install  libraries 

if(!require(tidyverse))  install.packages('tidyverse', repos = 'http://cran.us.r-project.org')
if(!require(caret))      install.packages('caret', repos = 'http://cran.us.r-project.org')
if(!require(data.table)) install.packages('data.table', repos = 'http://cran.us.r-project.org')
if(!require(dplyr))      install.packages('dplyr', repos = 'http://cran.us.r-project.org')
if(!require(ggplot2))    install.packages('ggplot2', repos = 'http://cran.us.r-project.org')
if(!require(hexbin))     install.packages('hexbin', repos = 'http://cran.us.r-project.org')

#Loading libraries 
library(tidyverse) 
library(caret)
library(data.table)
library(dplyr)
library(ggplot2)
```


```{r,load data, echo=FALSE, include=FALSE}
# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")

# if using R 3.6 or earlier:
#movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
#                                           title = as.character(title),
#                                           genres = as.character(genres))
# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                           title = as.character(title),
                                           genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

```

The function RMSE that it will used to observe effectiveness
$$ RMSE = \sqrt{\frac{1}{N}\displaystyle\sum_{u,i} (\hat{y}_{u,i}-y_{u,i})^{2}} $$
```{r,function RMSE, echo=TRUE, include=TRUE}
RMSE <- function(real_ratings = NULL, predicted_ratings = NULL) {
    sqrt(mean((real_ratings - predicted_ratings)^2))
}
```
The name of columns,type and description of columns in both datasets are six:

- **userId** ```<integer>```    Unique identification number for each user.
- **movieId** ```<numeric>```   Unique identification number for each movie.
- **rating** ```<numeric>```    Rating by one user of one movie . From 0.5 to 5 with 0.5 of increments.
- **timestamp** ```<integer>``` Timestamp for one specific rating provided by one user in seconds from 1970-01-01.
- **title** ```<character>```   title of each movie including the year of the release.
- **genres** ```<character>```  Genre of each movie separated by pipes.


The original dataset has been divided into 2 subsets "edx" and "validation" .

Dimension of movilens data sets "edx"   90 % of original dataset
```{r,Dimension edx, echo=FALSE, include=TRUE}
#edx
dim(edx)
```
Dimension of movilens data sets  "validation"  10 % of original dataset
```{r,Dimension validation, echo=FALSE, include=TRUE}
#validation
dim(validation)
```
A view of ten first rows of set edx
```{r,view data, echo=FALSE, include=TRUE}
#a view of ten first rows edx
head(edx,10)
```




Histogram of Rating distribution \

```{r,histogram rating, echo=FALSE, include=TRUE}
#histogram of rating to see distribution , integers are used more than fractions
edx%>%
  ggplot(. , aes(x=.$rating)) +
  geom_histogram(binwidth = 0.25,fill = "white",colour="blue",bins=10)+
  xlab("Rating") + 
  ylab("Count")+
  ggtitle("Rating Distribution ")+
  scale_y_continuous(breaks = c(seq(0, 3500000, 500000))) +
  theme_light()

```

How many users and movies in edx subset ?
```{r,user films, echo=FALSE, include=TRUE}
edx %>% summarize(Users = n_distinct(userId),
                  Movies = n_distinct(movieId)) 

```
Summary :  count,median , mean, max,1rd an 3rd quartiele by field 
The summary of the subsets "edx" and "validation" shows that there are no missing values.
Summary of **edx**
```{r,summary1, echo=FALSE, include=TRUE}
summary(edx)
```

Summary of **validation**
```{r,summary2, echo=FALSE, include=TRUE}
summary(validation)
```
another way ...
Does exists missing values in any column?  No,it does not
**edx**
```{r,missing1, echo=FALSE, include=TRUE}
sapply(edx, function(col) sum(is.na(col))) 
```
**validation**
```{r,missing2, echo=FALSE, include=TRUE}
sapply(validation, function(col) sum(is.na(col))) 
```

Users are not equally critical of their ratings, some users tend to give ratings 
much lower or higher than average. The next praphics includes only users who have 
rated at least 10 movies

```{r,ratings_by_users, echo=FALSE, include=TRUE,warning=FALSE}
edx %>%
  group_by(userId) %>%
  filter(n() >= 10) %>%
  summarize(avg_r = mean(rating)) %>%
  ggplot(aes(avg_r)) +
  geom_histogram(bins = 100, color = "green") +
  xlab("Mean rating") +
  ylab("Count of users") +
  ggtitle("Distribution of users by average of rating ") +
  scale_x_discrete(limits = c(seq(0.0,5,0.5))) +
  theme_light()
```
it looks like normal distribution with center close to 3.6

\newpage


# Dataset Pre-Processing 

The pre-processing phase is composed by this steps (always in both datasets):

1. Convert column timestamp to datetime format.
2. Extract the hour , day, month , the year and week day from the date.
3. Extract the year for each movie from the title.
4. Get antiquity.
5. Get number of rating per movie.
6. Separate each genre actually separated by pipe.
7. Get count genre per movie.
8. Convert columns to desidered data type.


Steps Development

1. Convert column timestamp to datetime format.
```{r,convert_datetime, echo=TRUE, include=TRUE}
edx$datetime        <- as.POSIXct(edx$timestamp, origin="1970-01-01")
validation$datetime <- as.POSIXct(validation$timestamp, origin="1970-01-01")
```


2. Extract the hour , day, month , the year and week day from the date in both datasets 
```{r,extract time, echo=TRUE, include=TRUE}
edx$weekday_Rate <- weekdays(edx$datetime)
edx$hour_Rate <- format(edx$datetime,"%H")
edx$year_Rate <- format(edx$datetime,"%Y")
edx$month_Rate <- format(edx$datetime,"%m")
validation$weekday_Rate <- weekdays(validation$datetime)
validation$hour_Rate <- format(validation$datetime,"%H")
validation$year_Rate <- format(validation$datetime,"%Y")
validation$month_Rate <- format(validation$datetime,"%m")
```
\newpage
\
3. Extract the year for each movie from the title.
```{r,year_movie, echo=TRUE, include=TRUE}
# edx 
edx <- edx %>%
   mutate(title = str_trim(title )) %>%
   extract(title,
           c("title_with_out_year", "year_movie"),
           regex = "^(.*) \\(([0-9 \\-]*)\\)$",
           remove = F) %>%
   mutate(release = if_else(str_length(year_movie) >= 5,
                            as.integer(str_split(year_movie, "-",simplify = T)[1]),
                            as.integer(year_movie))
   ) %>%
   mutate(title = if_else(is.na(title_with_out_year),
                          title,
                          title_with_out_year)
         ) %>%
  select(-title_with_out_year)
# validation 
validation <- validation %>%
   mutate(title = str_trim(title )) %>%
   extract(title,
           c("title_with_out_year", "year_movie"),
           regex = "^(.*) \\(([0-9 \\-]*)\\)$",
           remove = F) %>%
   mutate(release = if_else(str_length(year_movie) >= 5,
                            as.integer(str_split(year_movie, "-",simplify = T)[1]),
                            as.integer(year_movie))
   ) %>%
   mutate(title = if_else(is.na(title_with_out_year),
                          title,
                          title_with_out_year)
         ) %>%
  select(-title_with_out_year)
```

The graphics below shows the distribution by year of movie

```{r,year_movie2, echo=FALSE, include=TRUE,warning=FALSE}
edx%>%
  ggplot(. , aes(x=release,y=rating)) + 
  geom_bar(stat = "summary" ,fun = "mean",fill="khaki",colour="tan")

```

\newpage

4.Get antiquity 
\
The figure shows that old movies have more rating tan news , except the olders
\

```{r,get_antiquity0, echo=FALSE, include=TRUE}

edx<-edx %>%
   mutate(antiquity = as.numeric(year_Rate)-as.numeric(release)) 
validation<-validation %>%
   mutate(antiquity = as.numeric(year_Rate)-as.numeric(release)) 
```

```{r,get_antiquity1, echo=FALSE, include=TRUE}

edx %>%
   ggplot(. , aes(x=.$antiquity,y=.$rating)) +
   geom_bar(stat = "summary", fun = "mean",fill = "white", colour = "blue")+
   xlab("Antiquity of Movie") + 
   ylab("Average Rating")+
   ggtitle("Average Rating by  Antiquity of Movie") 

```
There are some ratings with a error, the rating year is lower than the year of the movie
**edx** dataset 175 rows
```{r,wrong_antiquity_edx, echo=FALSE, include=TRUE}
edx %>%
  filter(antiquity<0)%>%
  count()
```  
**validation** dataset 26 rows
```{r,wrong_antiquity_validation, echo=FALSE, include=TRUE}  
validation %>%
  filter(antiquity<0)%>%
  count()
```
Two options :
1 correct the year of rating and antiquity
2 drop the rows

option 2 is taken
```{r,drop_wrong_antiquity , echo=FALSE, include=TRUE}  
edx<-edx %>%
    filter(antiquity>=0)
validation<-validation %>%
    filter(antiquity>=0)  
```

\
In the next graphic : older movies have less ratings
\

```{r,get_antiquity2, echo=FALSE, include=TRUE}

edx %>%
   count(movieId,timestamp,userId,antiquity) %>%
   count(antiquity) %>%
   ggplot(aes(x=antiquity ,y=n)) +
   geom_bar(stat = "summary", fun = "sum",fill = "white", colour = "green")+
   xlab("Antiquity of Movie") + 
   ylab("Count of Rating")+
   ggtitle("Distribution of Rating by  Antiquity of Movie") 



```




\newpage
\
5. Get number of rating per movie
\
The first 10 movieId and "How many ratings they have?"
```{r,count_ratings, echo=FALSE, include=TRUE}
#aux df for calculation
df_edx_cnt_rating<-edx %>%
   group_by(movieId) %>%
   summarise(cnt_ratings=n())
head(df_edx_cnt_rating,10)   
```

\
More count of ratings mean High ratings per movies 

```{r,count_ratings2, echo=FALSE, include=TRUE}
edx<- edx %>% 
   left_join(df_edx_cnt_rating,by='movieId')
df_validation_cnt_rating<-edx %>%
   group_by(movieId) %>%
   summarise(cnt_ratings=n())
validation<- validation %>% 
   left_join(df_validation_cnt_rating,by='movieId')
edx %>%
   ggplot(. , aes(x=.$rating,y=.$cnt_ratings)) +
   geom_bar(stat = "summary", fun = "mean",fill="pink",colour="blue")+
   xlab("Rating") + 
   ylab("Rating per movie")+
   ggtitle("Average of Count of Rating per movie by  Rating value") 
```

\
More count of ratings are for less movies 

```{r,count_ratings3, echo=FALSE, include=TRUE}

edx%>%
  ggplot(. , aes(x=.$cnt_ratings)) + 
  geom_histogram(fill="blue",bins=100)+
  xlab("Count of Rating per movie") + 
  ylab("Count ")+
  ggtitle("Distribution of Count of Rating per movie")
```

\newpage
\
6.Get individual genre
\
Film noir have best average ratings as genre 
```{r,extract genre, echo=TRUE, include=TRUE, warning = FALSE}
# Extract the genre in edx datasets
edx <- edx %>%
   mutate(genre = fct_explicit_na(genres,na_level = "(No genres)")) %>%
   separate_rows(genre,sep = "\\|")
# Extract the genre in validation datasets
validation <- validation %>%
   mutate(genre = fct_explicit_na(genres,na_level = "(No genres)")) %>%
   separate_rows(genre,sep = "\\|")

```
```{r,extract genre2, echo=FALSE, include=TRUE, warning = FALSE}

edx %>% 
  ggplot(., aes(x=.$genre, y=.$rating,fill=.$genre)) +
  geom_boxplot()+stat_boxplot(geom = "errorbar",width = 0.5) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  theme(legend.position="none")+
  xlab("Genre") + 
  ylab("Rating")+
  ggtitle("Rating by Genre") 
```
```{r,genre3, echo=FALSE, include=TRUE, warning = FALSE}
edx %>% 
  ggplot(., aes(x=.$genre)) +
  geom_histogram(stat="count",fill = "white", colour = "blue")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  xlab("Genre") + 
  ylab("Count ")+
  ggtitle("Distribution by Genre")
```
\newpage
7. Get count genre per movie
The figure shows a little relation between the rating and the number of genres per movie\

```{r,count_genres, echo=FALSE, include=TRUE, warning=FALSE,message=FALSE}
# Extract the count of genre in edx datasets
#aux df for calculation
df_edx_count_genres<-edx %>%
   group_by(movieId,userId,datetime) %>%
   summarise(count_genres=n())
edx<- edx %>% 
   left_join(df_edx_count_genres,by=c("movieId" = "movieId", "userId" = "userId", "datetime" = "datetime"))
df_val_count_genres<-validation %>%
   group_by(movieId,userId,datetime) %>%
   summarise(count_genres=n())
validation<- validation %>% 
   left_join(df_val_count_genres,by=c("movieId" = "movieId", "userId" = "userId", "datetime" = "datetime"))


#edx<-edx%>%
#   mutate(count_genres=(sum(charToRaw(.$genres) == charToRaw("|") )+1) )
#validation<-validation%>%
#   mutate(count_genres=sum(charToRaw(.$genres) == charToRaw("|") ))
```

```{r,count_genres2, echo=FALSE, include=TRUE, warning=FALSE}
edx %>%
   ggplot(. , aes(x=.$rating,y=.$count_genres)) +
   geom_bar(stat = "summary", fun = "mean",fill = "white", colour = "red")+
   xlab("Rating")+ 
   ylab("Average Count of Genres per Movie")+
   ggtitle("Average Count of Genres per Movie by Rating") 

```
\newpage
8. Convert columns to desidered data type 
```{r,convert_numeric, echo=TRUE, include=TRUE}
edx$hour_Rate <- as.numeric(edx$hour_Rate)
edx$year_Rate <- as.numeric(edx$year_Rate)
edx$month_Rate <- as.numeric(edx$month_Rate)
edx$release <- as.numeric(edx$release)
edx$year_movie <- as.numeric(edx$year_movie)

validation$hour_Rate <- as.numeric(validation$hour_Rate)
validation$year_Rate <- as.numeric(validation$year_Rate)
validation$month_Rate <- as.numeric(validation$month_Rate)
validation$release <- as.numeric(validation$release)
validation$year_movie <- as.numeric(validation$year_movie)

```
Remove unnecessary columns on edx and validation dataset
```{r,remove_fields, echo=TRUE, include=TRUE}
edx <-edx%>% 
   select(userId, movieId, rating, title, genre, release, weekday_Rate, 
          hour_Rate, year_Rate, month_Rate,count_genres,antiquity,cnt_ratings)
validation <- validation %>% 
   select(userId, movieId, rating, title, genre, release, weekday_Rate, 
          hour_Rate, year_Rate, month_Rate,count_genres,antiquity,cnt_ratings)
```
\
Summary to see is there are NA values count, median,mean ,min,max,1rd an 3rd quartiele
```{r,summary , echo=FALSE, include=TRUE}
summary(edx)
```

\newpage

After preprocessing the data, **edx** dataset looks like this:

**Processed edx datadaset**

```{r,head edx, echo=FALSE, include=TRUE}
# Output the processed dataset
head(edx) 
```




# Model Building and Evaluation

## Naive Baseline Model

Naive Model predict the average of all movie ratings is approximately 3.53


```{r,Naive model mean , echo=FALSE, include=TRUE}
paste("The mean of rating in edx set is:", as.character(mean(edx$rating)))

```

## Model 1 : Naive Mean-Baseline 

The formula used for this model:

$$Y_{u,i} = \hat{\mu} + \varepsilon_{u,i}$$

With $\hat{\mu}$ is the mean and $\varepsilon_{i,u}$ is the independent sampled errors.

```{r,rmse naive model}
mu_hat <- mean(edx$rating)
rmse_model_1 <- RMSE(validation$rating, mu_hat)
c_model<-"Model 1: Naive Mean Baseline  "
results <- data.frame(model=c_model, RMSE=rmse_model_1)
tibble(Method = c_model, RMSE = rmse_model_1)
```

## Model 2 : Movie-Based Model, a Content-based Approach

The first Non-Naive Model takes into account the content. In this case the movies that are rated higher or lower resperct to each other.

The formula used is:

$$Y_{u,i} = \hat{\mu} + b_i + \epsilon_{u,i}$$

With $\hat{\mu}$ is the mean and $\varepsilon_{i,u}$ is the independent errors sampled from the same distribution centered at 0. The $b_i$ is a measure for the popularity of movie $i$, i.e. the bias of movie $i$.

```{r,movie base model, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Calculate the average by movie
movie_avgs <- edx %>%
   group_by(movieId) %>%
   summarize(b_i = mean(rating - mu_hat))
# Graph the Movie Bias
movie_avgs %>% ggplot(aes(b_i)) +
  geom_histogram(color = "black", fill = "pink", bins = 100) +
  xlab("Movie Bias") +
  ylab("Count ") +
  theme_bw()

# Compute the predicted ratings on validation dataset
rmse_movie_model <- validation %>%
   left_join(movie_avgs, by='movieId') %>%
   mutate(pred = mu_hat + b_i) %>%
   pull(pred)
rmse_model_2 <- RMSE(validation$rating, rmse_movie_model)
# Adding results to the results dataset
c_model<-"Model 2: Movie-Based Model  "
results <- results %>% add_row(model=c_model, RMSE=rmse_model_2)
tibble(Method = c_model, RMSE = rmse_model_2)
```

The RMSE on the **validation** dataset is **0.94107**. It better than the Naive Mean-Baseline Model, but it is also very far from the target RMSE (below 0.87) and that indicates poor performance for the model.

## Model 3: Movie + User Model, a User-based approach

The second Non-Naive Model consider that the users have different tastes and rate differently.

The formula used is:

$$Y_{u,i} = \hat{\mu} + b_i + b_u + \epsilon_{u,i}$$

With $\hat{\mu}$ is the mean and $\varepsilon_{i,u}$ is the independent errors sampled from the same distribution centered at 0. The $b_i$ is a measure for the popularity of movie $i$, i.e. the bias of movie $i$. The  $b_u$ is a measure for the mildness of user $u$, i.e. the bias of user $u$.

```{r,movie_user_based_model, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Calculate the average by movie
movie_avgs <- edx %>%
   group_by(movieId) %>%
   summarize(b_i = mean(rating - mu_hat))
# Calculate the average by user
user_avgs <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   group_by(userId) %>%
   summarize(b_u = mean(rating - mu_hat - b_i))
# Graph the User Bias
user_avgs %>% ggplot(aes(b_u)) +
  geom_histogram(color = "black", fill = "red", bins = 10) +
  xlab("User Bias") +
  ylab("Count ") +
  theme_bw()

# Compute the predicted ratings on validation dataset
rmse_movie_user_model <- validation %>%
   left_join(movie_avgs, by='movieId') %>%
   left_join(user_avgs, by='userId') %>%
   mutate(pred = mu_hat + b_i + b_u) %>%
   pull(pred)
rmse_model_3 <- RMSE(validation$rating, rmse_movie_user_model)
# Adding the results to the results dataset
c_model<-"Model 3: Movie+User Based Model  "
results <- results %>% add_row(model=c_model, RMSE=rmse_model_3)
tibble(Method = c_model, RMSE = rmse_model_3)
```

The RMSE on the **validation** dataset is **0.86337** and this is very good. The Movie+User Based Model reaches the desidered performance but applying the regularization techniques, can improve the performance just a little.

## Model 4 : Movie + User + Genre Model, the Genre Popularity

The formula used is:

$$Y_{u,i} = \hat{\mu} + b_i + b_u + b_{u,g} + \epsilon_{u,i}$$

With $\hat{\mu}$ is the mean and $\varepsilon_{i,u}$ is the independent errors sampled from the same distribution centered at 0. The $b_i$ is a measure for the popularity of movie $i$, i.e. the bias of movie $i$. The  $b_u$ is a measure for the mildness of user $u$, i.e. the bias of user $u$. The  $b_{u,g}$ is a measure for how much a user $u$ likes the genre $g$.

```{r,movie_user_genre_model, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Calculate the average by movie
movie_avgs <- edx %>%
   group_by(movieId) %>%
   summarize(b_i = mean(rating - mu_hat))
# Calculate the average by user
user_avgs <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   group_by(userId) %>%
   summarize(b_u = mean(rating - mu_hat - b_i))
genre_pop <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   left_join(user_avgs, by='userId') %>%
   group_by(genre) %>%
   summarize(b_u_g = mean(rating - mu_hat - b_i - b_u))

# Graph the Genre Popularity Bias
genre_pop %>% ggplot(aes(b_u_g)) +
  geom_histogram(color = "black", fill = "Green", bins = 100) +
  xlab("Genre Bias") +
  ylab("Count ") +
  theme_bw()

# Compute the predicted ratings on validation dataset
rmse_movie_user_genre_model <- validation %>%
   left_join(movie_avgs, by='movieId') %>%
   left_join(user_avgs, by='userId') %>%
   left_join(genre_pop, by='genre') %>%
   mutate(pred = mu_hat + b_i + b_u + b_u_g) %>%
   pull(pred)
rmse_model_4 <- RMSE(validation$rating, rmse_movie_user_genre_model)
# Adding the results to the results dataset
c_model<-"Model 4:Movie+User+Genre Based Model  "
results <- results %>% add_row(model=c_model, RMSE=rmse_model_4)
tibble(Method = c_model, RMSE = rmse_model_4)
```

The RMSE on the **validation** dataset is **0.86327** and this is very good.  The Movie+User+Genre Based Model reaches the desidered performance but adding the ```genre``` predictor, doesn't improve significantly the model's performance. Applying the regularization techniques, can improve the performance just a little.



## Model 5:Movie + User + Genre + Antiquity Model, the Antiquity  Popularity

The formula used is:

$$Y_{u,i} = \hat{\mu} + b_i + b_u + b_{u,g}  + b_a + \epsilon_{u,i}$$



With $\hat{\mu}$ is the mean and $\varepsilon_{i,u}$ is the independent errors sampled from the same distribution centered at 0. 
The $b_i$ is a measure for the popularity of movie $i$, i.e. the bias of movie $i$. 
The  $b_u$ is a measure for the mildness of user $u$, i.e. the bias of user $u$. 
The  $b_{u,g}$ is a measure for how much a user $u$ likes the genre $g$.
The  $b_a$ is a measure for the antiquity of movie $i$, i.e. the bias of antiquity of movie $i$. 

```{r,plus_antiquity, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Calculate the average by movie
movie_avgs <- edx %>%
   group_by(movieId) %>%
   summarize(b_i = mean(rating - mu_hat))
# Calculate the average by user
user_avgs <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   group_by(userId) %>%
   summarize(b_u = mean(rating - mu_hat - b_i))

genre_pop <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   left_join(user_avgs, by='userId') %>%
   group_by(genre) %>%
   summarize(b_u_g = mean(rating - mu_hat - b_i - b_u))

antiquity_avg <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   left_join(user_avgs, by='userId') %>%
   left_join(genre_pop, by='genre') %>%
   group_by(antiquity) %>%
   summarize(b_a = mean(rating - mu_hat - b_i - b_u - b_u_g))

# Graph the antiquity Popularity Bias
antiquity_avg %>% ggplot(aes(b_a)) +
  geom_histogram(color = "black", fill = "deepskyblue", bins = 100) +
  xlab("Antiquity Bias") +
  ylab("Count ") +
  theme_bw()



# Compute the predicted ratings on validation dataset
rmse_movie_user_genre_antiquity_model <- validation %>%
   left_join(movie_avgs,    by='movieId') %>%
   left_join(user_avgs,     by='userId') %>%
   left_join(genre_pop,     by='genre') %>%
   left_join(antiquity_avg, by='antiquity') %>%
   mutate(pred = mu_hat + b_i + b_u + b_u_g + b_a) %>%
   pull(pred)
rmse_model_5 <- RMSE(validation$rating, rmse_movie_user_genre_antiquity_model)
# Adding the results to the results dataset
c_model<-"Movie+User+Genre+Antiquity Based Model  "
results <- results %>% add_row(model=c_model, RMSE=rmse_model_5)
tibble(Method = c_model, RMSE = rmse_model_5)
```

The RMSE on the **validation** dataset is **0.86274** and this is very good.  The Movie+User+Genre+Antiquity Based Model reaches the desidered performance but adding the ```Antiquity``` predictor, doesn't improve significantly the model's performance. Applying the regularization techniques, can improve the performance just a little.



## Model 6: Movie + User + Genre + Antiquity +Count Rating Model

The formula used is:

$$Y_{u,i} = \hat{\mu} + b_i + b_u + b_{u,g}  + b_a +  b_c + \epsilon_{u,i}$$





With $\hat{\mu}$ is the mean and $\varepsilon_{i,u}$ is the independent errors sampled from the same distribution centered at 0. 
The $b_i$ is a measure for the popularity of movie $i$, i.e. the bias of movie $i$. 
The  $b_u$ is a measure for the mildness of user $u$, i.e. the bias of user $u$. 
The  $b_{u,g}$ is a measure for how much a user $u$ likes the genre $g$.
The  $b_a$ is a measure for the antiquity of movie $i$, i.e. the bias of antiquity of movie $i$. 
The  $b_c$ is a measure for the count of rating of movie $i$, i.e. the bias of count of rating of movie $i$. 

```{r,b_a, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Calculate the average by movie
movie_avgs <- edx %>%
   group_by(movieId) %>%
   summarize(b_i = mean(rating - mu_hat))
# Calculate the average by user
user_avgs <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   group_by(userId) %>%
   summarize(b_u = mean(rating - mu_hat - b_i))

genre_pop <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   left_join(user_avgs, by='userId') %>%
   group_by(genre) %>%
   summarize(b_u_g = mean(rating - mu_hat - b_i - b_u))

antiquity_avg <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   left_join(user_avgs, by='userId') %>%
   left_join(genre_pop, by='genre') %>%
   group_by(antiquity) %>%
   summarize(b_a = mean(rating - mu_hat - b_i - b_u - b_u_g))

cant_rating_avg <- edx %>%
   left_join(movie_avgs, by='movieId') %>%
   left_join(user_avgs, by='userId') %>%
   left_join(genre_pop, by='genre') %>%
   left_join(antiquity_avg, by='antiquity') %>%
   group_by(cnt_ratings) %>%
   summarize(b_c = mean(rating - mu_hat - b_i - b_u - b_u_g - b_a ))
# Graph the count rating  Bias
cant_rating_avg %>% ggplot(aes(b_c)) +
  geom_histogram(color = "black", fill = "yellow", bins = 100) +
  xlab("Count Rating Bias") +
  ylab("Count ") +
  theme_bw()
# Compute the predicted ratings on validation dataset
rmse_movie_user_genre_antiquity_cant_rating_model <- validation %>%
   left_join(movie_avgs,      by='movieId') %>%
   left_join(user_avgs,       by='userId') %>%
   left_join(genre_pop,       by='genre') %>%
   left_join(antiquity_avg,   by='antiquity') %>%
   left_join(cant_rating_avg, by='cnt_ratings') %>%
   mutate(pred = mu_hat + b_i + b_u + b_u_g + b_a + b_c) %>%
   pull(pred)
rmse_model_6 <- RMSE(validation$rating, rmse_movie_user_genre_antiquity_cant_rating_model)
# Adding the results to the results dataset
c_model<-"Model 6: Movie + User + Genre + Antiquity +Count Rating Model  "
results <- results %>% 
   add_row(model=c_model,RMSE=rmse_model_6)
tibble(Method = c_model, RMSE = rmse_model_6)
```

The RMSE on the **validation** dataset is **0.86187** and this is very good.  The Movie+User+Genre+Antiquity Based Model reaches the desidered performance but adding the ```Antiquity``` predictor, doesn't improve significantly the model's performance. Applying the regularization techniques, can improve the performance just a little.



## Regularization

It allows to add a penalty $\lambda$ (lambda) to penalizes movies with large estimates from a small sample size. In order to optimize $b_i$, it necessary to use this equation:

$$\frac{1}{N} \sum_{u,i} (y_{u,i} - \mu - b_{i})^{2} + \lambda \sum_{i} b_{i}^2$$   

reduced to this equation:   

$$\hat{b_{i}} (\lambda) = \frac{1}{\lambda + n_{i}} \sum_{u=1}^{n_{i}} (Y_{u,i} - \hat{\mu}) $$  

### Regularized Movie-Based Model

```{r,regularized_movie_based_model, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Define a table of lambdas
lambdas <- seq(0, 10, 0.1)
# Compute the predicted ratings on validation dataset using different values of lambda
rmses <- sapply(lambdas, function(lambda) {

   
  # Calculate the average by user
  
   b_i <- edx %>%
      group_by(movieId) %>%
      summarize(b_i = sum(rating - mu_hat) / (n() + lambda))
   
   # Compute the predicted ratings on validation dataset
   
   predicted_ratings <- validation %>%
      left_join(b_i, by='movieId') %>%
      mutate(pred = mu_hat + b_i) %>%
      pull(pred)
   
   # Predict the RMSE on the validation set
   
   return(RMSE(validation$rating, predicted_ratings))
})
# plot the result of lambdas
df <- data.frame(RMSE = rmses, lambdas = lambdas)
ggplot(df, aes(lambdas, rmses)) +
   theme_classic()  +
   geom_point() +
   labs(title = "RMSEs vs Lambdas - Regularized Movie Based Model",
        y = "RMSEs",
        x = "lambdas")
# Get the lambda value that minimize the RMSE
min_lambda <- lambdas[which.min(rmses)]
# Predict the RMSE on the validation set
rmse_reg_model_2 <- min(rmses)
# Adding the results to the results dataset
c_model<-"Regularized Movie-Based Model"
results <- results %>% add_row(model=c_model, RMSE=rmse_reg_model_2)
tibble(Method = c_model, RMSE = rmse_reg_model_2)
```

The RMSE on the **validation** dataset is **0.94103** and this is better than without Regularized. The Movie Based Model is not reaches yet the desired performance but applying the regularization techniques improve it just a little.

### Regularized Movie+User Model

```{r,regularized_movie_user_model, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Define a table of lambdas
lambdas <- seq(0, 25, 0.1)
# Compute the predicted ratings on validation dataset using different values of lambda
rmses <- sapply(lambdas, function(lambda) {

   # Calculate the average by user
   
   b_i <- edx %>%
      group_by(movieId) %>%
      summarize(b_i = sum(rating - mu_hat) / (n() + lambda))
   
   # Calculate the average by user
   
   b_u <- edx %>%
      left_join(b_i, by='movieId') %>%
      group_by(userId) %>%
      summarize(b_u = sum(rating - b_i - mu_hat) / (n() + lambda))
   
   # Compute the predicted ratings on validation dataset
   
   predicted_ratings <- validation %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      mutate(pred = mu_hat + b_i + b_u) %>%
      pull(pred)
   
   # Predict the RMSE on the validation set
   
   return(RMSE(validation$rating, predicted_ratings))
})
# plot the result of lambdas
df <- data.frame(RMSE = rmses, lambdas = lambdas)
ggplot(df, aes(lambdas, rmses)) +
   theme_classic()  +
   geom_point() +
   labs(title = "RMSEs vs Lambdas - Regularized Movie+User Model",
        y = "RMSEs",
        x = "lambdas")
# Get the lambda value that minimize the RMSE
min_lambda <- lambdas[which.min(rmses)]
# Predict the RMSE on the validation set
rmse_reg_model_3 <- min(rmses)
# Adding the results to the results dataset
c_model<-"Regularized Movie+User Based Model"
results <- results %>% add_row(model=c_model, RMSE=rmse_reg_model_3)
tibble(Method = c_model, RMSE = rmse_reg_model_3)
```

The RMSE on the **validation** dataset is **0.86275**. The Regularized Movie+User Based Model improves just a little the result of the Non-Regularized Model.

### Regularized Movie+User+Genre Model

```{r,regularized_movie_user_genre_model, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Define a table of lambdas
lambdas <- seq(0, 25, 0.1)
# Compute the predicted ratings on validation dataset using different values of lambda
rmses <- sapply(lambdas, function(lambda) {

   # Calculate the average by user
   
   b_i <- edx %>%
      group_by(movieId) %>%
      summarize(b_i = sum(rating - mu_hat) / (n() + lambda))
   
   # Calculate the average by user
   
   b_u <- edx %>%
      left_join(b_i, by='movieId') %>%
      group_by(userId) %>%
      summarize(b_u = sum(rating - b_i - mu_hat) / (n() + lambda))
   
    b_u_g <- edx %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      group_by(genre) %>%
      summarize(b_u_g = sum(rating - b_i - mu_hat - b_u) / (n() + lambda))
   
   # Compute the predicted ratings on validation dataset
   
   predicted_ratings <- validation %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      left_join(b_u_g, by='genre') %>%
      mutate(pred = mu_hat + b_i + b_u + b_u_g) %>%
      pull(pred)
   
   # Predict the RMSE on the validation set
   
   return(RMSE(validation$rating, predicted_ratings))
})
# plot the result of lambdas
df <- data.frame(RMSE = rmses, lambdas = lambdas)
ggplot(df, aes(lambdas, rmses)) +
   theme_classic()  +
   geom_point() +
   labs(title = "RMSEs vs Lambdas - Regularized Movie+User+Genre Model",
        y = "RMSEs",
        x = "lambdas")
# Get the lambda value that minimize the RMSE
min_lambda <- lambdas[which.min(rmses)]
# Predict the RMSE on the validation set
rmse_reg_model_4 <- min(rmses)
# Adding the results to the results dataset
c_model<-"Regularized Movie+User+Genre Based Model "
results <- results %>% add_row(model=c_model, RMSE=rmse_reg_model_4)
tibble(Method = c_model, RMSE = rmse_reg_model_4)
```

The RMSE on the **validation** dataset is **0.86267** . 
The Regularized Movie+User+Genre Based Model improves just a little the result of the Non-Regularized Model. 
As the Non-Regularized Model, the ```genre``` predictor doesn't improve significantly the model's performance.


### Regularized Movie+User+Genre+Antiquity Model

```{r,regularized_movie_user_genre_antiquity_model, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Define a table of lambdas
lambdas <- seq(0, 25, 0.1)
# Compute the predicted ratings on validation dataset using different values of lambda
rmses <- sapply(lambdas, function(lambda) {
   # Calculate the average by user
   
   b_i <- edx %>%
      group_by(movieId) %>%
      summarize(b_i = sum(rating - mu_hat) / (n() + lambda))
   
   # Calculate the average by user
   
   b_u <- edx %>%
      left_join(b_i, by='movieId') %>%
      group_by(userId) %>%
      summarize(b_u = sum(rating - b_i - mu_hat) / (n() + lambda))
   
   b_u_g <- edx %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      group_by(genre) %>%
      summarize(b_u_g = sum(rating - b_i - mu_hat - b_u) / (n() + lambda))
   
   b_a <- edx %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      left_join(b_u_g, by='genre') %>%
      group_by(antiquity) %>%
      summarize(b_a = sum(rating - b_i - mu_hat - b_u - b_u_g ) / (n() + lambda))
   
   # Compute the predicted ratings on validation dataset
   
   predicted_ratings <- validation %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      left_join(b_u_g, by='genre') %>%
      left_join(b_a, by='antiquity') %>%
      mutate(pred = mu_hat + b_i + b_u + b_u_g + b_a) %>%
      pull(pred)
   
   # Predict the RMSE on the validation set
   
   return(RMSE(validation$rating, predicted_ratings))
})
# plot the result of lambdas
title_gg<-"RMSEs vs Lambdas - Regularized Movie+User+Genre+Antiquity Model"
df <- data.frame(RMSE = rmses, lambdas = lambdas)
ggplot(df, aes(lambdas, rmses)) +
   theme_classic()  +
   geom_point() +
   labs(title = title_gg,
        y = "RMSEs",
        x = "lambdas")
# Get the lambda value that minimize the RMSE
min_lambda <- lambdas[which.min(rmses)]
# Predict the RMSE on the validation set
rmse_reg_model_5 <- min(rmses)
# Adding the results to the results dataset
c_model<-"Regularized Movie+User+Genre+Antiquity Based Model"
results <- results %>% 
  add_row(model=c_model,RMSE=rmse_reg_model_5)
tibble(Method = c_model, RMSE = rmse_reg_model_5)
```

The RMSE on the **validation** dataset is **0.86211**. 
The Regularized Movie+User+Genre+Antiquity Base Model improves just a little the result of the Non-Regularized Model. 
As the Non-Regularized Model, the ```Antiquity``` predictor doesn't improve significantly the model's performance.



### Regularized Movie+User+Genre+Antiquity+count Rating Model

```{r,regularized_movie_user_genre_antiquity_count_rating_model, echo=TRUE, include=TRUE}
# Calculate the average of all movies
mu_hat <- mean(edx$rating)
# Define a table of lambdas
lambdas <- seq(0, 25, 0.1)
# Compute the predicted ratings on validation dataset using different values of lambda
rmses <- sapply(lambdas, function(lambda) {

   # Calculate the average by user
   
   b_i <- edx %>%
      group_by(movieId) %>%
      summarize(b_i = sum(rating - mu_hat) / (n() + lambda))
   
   # Calculate the average by user
   
   b_u <- edx %>%
      left_join(b_i, by='movieId') %>%
      group_by(userId) %>%
      summarize(b_u = sum(rating - b_i - mu_hat) / (n() + lambda))
   
   b_u_g <- edx %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      group_by(genre) %>%
      summarize(b_u_g = sum(rating - b_i - mu_hat - b_u) / (n() + lambda))
   
   b_a <- edx %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      left_join(b_u_g, by='genre') %>%
      group_by(antiquity) %>%
      summarize(b_a = sum(rating - b_i - mu_hat - b_u - b_u_g ) / (n() + lambda))
   
   b_c <- edx %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      left_join(b_u_g, by='genre') %>%
      left_join(b_a, by='antiquity') %>%
      group_by(cnt_ratings) %>%
      summarize(b_c = sum(rating - b_i - mu_hat - b_u - b_u_g - b_a) / (n() + lambda))

      # Compute the predicted ratings on validation dataset
   
   predicted_ratings <- validation %>%
      left_join(b_i, by='movieId') %>%
      left_join(b_u, by='userId') %>%
      left_join(b_u_g, by='genre') %>%
      left_join(b_a, by='antiquity') %>%
      left_join(b_c, by='cnt_ratings') %>%
      mutate(pred = mu_hat + b_i + b_u + b_u_g + b_a + b_c) %>%
      pull(pred)
   
   # Predict the RMSE on the validation set
   
   return(RMSE(validation$rating, predicted_ratings))
})
# plot the result of lambdas
title_gg<-"RMSEs vs Lambdas - Regularized Movie+User+Genre+Antiquity+Cant Rating Model"
df <- data.frame(RMSE = rmses, lambdas = lambdas)
ggplot(df, aes(lambdas, rmses)) +
   theme_classic()  +
   geom_point() +
   labs(title = title_gg,
        y = "RMSEs",
        x = "lambdas")
# Get the lambda value that minimize the RMSE
min_lambda <- lambdas[which.min(rmses)]
# Predict the RMSE on the validation set
rmse_reg_model_6 <- min(rmses)
# Adding the results to the results dataset
c_model<-"Regularized Movie+User+Genre+Antiquity+Count Rating Based Model"
results <- results %>% 
  add_row(model=c_model, RMSE=rmse_reg_model_6)
tibble(Method = c_model, RMSE = rmse_reg_model_6)
```

The RMSE on the **validation** dataset is **0.86135** and this is the best result of the builted models. 
The Regularized Movie+User+Genre+Antiquity+count Rating  Model improves just a little the result of the Non-Regularized Model. As in the Non-Regularized Model, the **count Rating** predictor doesn't improve significantly the model's performance.


# Additional notes
The kable function was used at the begginig, but when pdf was generated it does work (using MikTex in Windows 10).
weekday_Rate was not use in any model because that was indifferent.
hour_Rate was not use in any model because that was indifferent.
year_Rate was not use in any model because that was indifferent except 1995 when the avg was 4, but for future It will does not make sense.


# Results

This is the summary results for all the model builted, trained on **edx** dataset and validated on the **validation** dataset.

```{r, echo=TRUE, include=TRUE}
# Shows the results
results 
```
